---
title: "Create patches"
format: html
editor: visual
---

## Introduction

This document introduces the first step of the analysis: defining spatial units for downstream testing. The main challenge is that “patches” do not have to be fixed in advance, so the pipeline should stay flexible enough to accept algorithmic patches, manually drawn regions, domains, or cell neighborhoods.

::: callout-important
**Goals**

-   Build a patch-definition workflow that can use different inputs, including patchDE, SVA, domains, or cell neighborhoods.

-   Compare alternative patching strategies on the two planned datasets, MERFISH MouseColonIBD and CosMx colon cancer.

-   Produce patch-level objects that are ready for differential expression analysis in the next document.
:::

## Data Preparation

We use the `MouseColonIbdCadinu2024` dataset from the `MerfishData` package. This dataset contains MERFISH data from a mouse model of colon inflammation (DSS treatment).

```{r}
library(MerfishData)
library(SpatialExperiment)
library(InSituCor)
source("patchDE.R")

# Load the dataset
spe <- MerfishData::MouseColonIbdCadinu2024()
colnames(spe) <- paste0("Cell", seq_len(ncol(spe)))

# Choose sample 5 slice 1 (3 days of DSS treatment)
spe_sub <- spe[, spe$sample_id == 5 & spe$slice_id == 1]

# Get spatial coordinates (convert to mm for patch definition)
xy <- as.matrix(spatialCoords(spe_sub))
xy_mm <- xy * 0.000109 # 1 pixel ~ 109 nm
rownames(xy_mm) <- colnames(spe_sub)
```


## Defining Exposure

Our biological question is: **How do epithelial cells behave in proximity of immune cells?** We define the "exposure" as the number of immune cells in the local neighborhood of each epithelial cell.

```{r}
# Get spatial neighbors matrix (N=50)
neighbors <- InSituCor:::nearestNeighborGraph(x = xy[,1], y = xy[,2], N = 50, subset = 1)

# Count neighboring cell types (tier1 classification)
env <- InSituCor:::neighbor_tabulate(as.vector(spe_sub$tier1), neighbors) 
rownames(env) <- rownames(xy_mm)

# Focus on Epithelial cells
is_epithelial <- spe_sub$tier1 == "Epithelial"
is_MU4 <- spe_sub$leiden_neigh == "MU4"
X <- env[is_epithelial & is_MU4, "Immune"]
```

## Patch Creation

We create patches within a specific neighborhood (e.g., "MU4") to control for larger-scale tissue structure, or across all epithelial cells.

```{r}
# Example: Create patches across all epithelial cells
n_patches <- 100
patches <- getPatches(
    xy = xy_mm[is_epithelial & is_MU4, ],
    X = X,
    npatches = n_patches, 
    maxradius = 0.1, 
    bitesize = 0.1,
    effectivezerodist = 0.001,
    plotprogress = FALSE)

# Store patch assignments in the SpatialExperiment object
spe_sub$patch <- NA
spe_sub$patch[is_epithelial & is_MU4] <- patches
```


## Visualization

Visualizing the patches helps confirm they are spatially contiguous and capture regions with varying immune cell exposure.

```{r}
# Define patch polygons for plotting
polys <- list()
use_cells <- which(is_epithelial & !is.na(spe_sub$patch))
for (p_id in setdiff(unique(spe_sub$patch), NA)) {
    inds <- which(spe_sub$patch[use_cells] == p_id)
    if(length(inds) > 3) {
        polys[[p_id]] <- xy_mm[use_cells, ][inds, ][chull(xy_mm[use_cells, ][inds, ]), ]
    }
}

# Plot patches
plot(xy_mm[is_epithelial, ], pch = 16, cex = 0.5, col = "grey80", main = "Epithelial Patches")
points(xy_mm[use_cells, ], pch = 16, cex = 0.5, 
       col = rainbow(n_patches)[as.numeric(as.factor(spe_sub$patch[use_cells]))])
for (p_id in names(polys)) {
    polygon(polys[[p_id]], border = "black")
}
```

