---
title: "Spatially stratified differential expression"
format: html
editor: visual
---

## Introduction

This document focuses on differential expression within patches while accounting for the fact that nearby cells are correlated. The key question is whether spatially aware methods give more reliable or more biologically meaningful results than standard non-spatial analyses.

::: callout-important
**Goals**

-   Test whether spatial autocorrelation matters in small patches and decide when it must be modeled explicitly.

-   Compare spatially aware and non-spatial DE methods on the same patches.

-   Evaluate whether spatial correction changes the biological interpretation, for example in tumor cells near T cells or epithelial cells near T cells.

-   Check residual spatial autocorrelation after modeling to see whether the spatial structure has really been removed.
:::

## Normalization and Data Subset

Prior to differential expression analysis, we normalize the raw counts. We focus on epithelial cells that are part of a defined patch.

```{r}
library(Matrix)
source("patchDE.R")

# Use logcounts from SpatialExperiment object if available, otherwise normalize
# counts <- assay(spe_sub, "counts")
# norm <- log1p(sweep(counts, 2, colSums(counts)/1e4, "*"))
norm <- assay(spe_sub, "logcounts")

# Subset for epithelial cells in patches
is_epithelial <- spe_sub$tier1 == "Epithelial"
is_MU4 <- spe_sub$leiden_neigh == "MU4"
in_patch <- !is.na(spe_sub$patch)
use <- is_epithelial & is_MU4 & in_patch

# Logcounts matrix (cells x genes) for patchDE
y <- t(as.matrix(norm[, use]))
```


## Patch-wise Differential Expression

We perform differential expression analysis within each patch using the `patchDE` function. This function uses a fast linear regression approach (`hastyDE`) across each patch.

```{r}
# Define design matrix (e.g., exposure variable)
df <- data.frame(Immune_Exposure = env[use, "Immune"])
patch_assignments <- spe_sub$patch[use]

# Run DE across all patches
de_results <- patchDE(y = y, 
                      df = df, 
                      patch = patch_assignments)

# Example: Result for a specific gene across all patches
gene_interest <- "Tnf"
res_Tnf <- data.frame(
    est = de_results$Immune_Exposure$ests[gene_interest, ],
    p = de_results$Immune_Exposure$pvals[gene_interest, ],
    se = de_results$Immune_Exposure$ses[gene_interest, ]
)
```


## Spatially Aware Modeling (Future Work)

The `patchDE` function currently uses a simple linear model within each patch. In more advanced analyses, we can compare these results with models that explicitly include a spatial correlation structure (e.g., Gaussian Process) to determine if local autocorrelation still influences the results after patching.

``` r
# Note: Further investigation into residual spatial autocorrelation 
# can be performed using Moran's I on model residuals.
```
