---
title: "meta-analyses"
format: html
editor: visual
---

## **Introduction**

This document summarizes the patch-level DE results and asks whether some groups of patches show a consistent or enriched response. The aim is to move from patch-wise results to higher-level biological patterns across patches and datasets.

::: callout-important
**Goals.**

-   Identify subgroups of patches enriched for a common DE relationship.

-   Choose and justify meta-analysis strategies that combine evidence across patches.

-   Compare results across the two datasets to see whether the same spatial relationships recur in both systems.

-   Produce a final synthesis that links patch structure, spatial DE, and biological interpretation.
:::

## Meta-analysis of Patch Results

We combine patch-level DE estimates using inverse-variance weighting. This provides a robust global summary of the effects across the entire tissue sample.

```{r}
# Extract results for the variable of interest
res <- de_results$Immune_Exposure

# Use inverse-variance weighting to combine results across patches
# Calculate meta-estimates and their standard errors
# Handle cases where SE might be 0 or NA
metawts <- pmax(res$ses, 0.1)^-1/2
metaest <- rowSums(res$ests * metawts, na.rm = TRUE) / rowSums(metawts, na.rm = TRUE)
metaSE <- sqrt(1 / rowSums(metawts))

# Calculate Z-scores and identify consistent genes
metaZ <- metaest / metaSE
is_consistent <- abs(metaZ) > 2

# Summary of top consistent genes
top_genes <- head(sort(abs(metaZ[is_consistent]), decreasing = TRUE))
```

## Cross-dataset Comparison

By repeating this pipeline on both the CosMx colon cancer and MERFISH MouseColonIBD datasets, we can evaluate whether the response of cells (epithelial or tumor) to immune cell proximity is consistent across different systems and technologies.

```r
# Example: Overlay meta-results from two datasets to identify common spatial programs
# common_genes <- intersect(consistent_genes_merfish, consistent_genes_cosmx)
```

## Visualization of Results

Visualizing the meta-analysis results through heatmaps or volcano plots helps summarize the findings across patches.

### Global Meta-analysis Summary
A volcano plot of the meta-estimates against the Z-scores identifies genes with the most consistent spatial response across the entire sample.

```{r}
library(ggplot2)

meta_df <- data.frame(
  gene = names(metaest),
  estimate = metaest,
  zscore = metaZ,
  consistent = is_consistent
)

ggplot(meta_df, aes(x = estimate, y = abs(zscore), color = consistent)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Meta-analysis Volcano Plot",
       x = "Average Effect Size (Meta-estimate)",
       y = "|Z-score|") +
  scale_color_manual(values = c("grey", "red"))
```

### Patch-level Consistency
We use a heatmap to inspect the t-statistics of the most consistent genes across all patches. This helps identify if certain patches are driving the global signal or if the effect is truly widespread.

```{r}
library(pheatmap)

# Calculate t-statistics per patch
t_stats <- res$ests / res$ses
t_stats[is.na(t_stats)] <- 0

# Select top consistent genes for the heatmap
top_consistent_genes <- names(sort(abs(metaZ), decreasing = TRUE))[1:30]

pheatmap(t_stats[top_consistent_genes, ], 
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-5, 5, length.out = 101),
         main = "Patch-level t-statistics (Top 30 Genes)",
         show_colnames = FALSE)
```

### Specific Gene Inspection
Finally, we can map the expression of a specific gene of interest alongside the patch boundaries and the exposure variable (Immune proximity).

```{r}
# Example gene of interest
goi <- "Egr1"

# Create a plotting data frame for cells in patches
plot_df <- data.frame(
  x = xy_mm[use, 1],
  y = xy_mm[use, 2],
  expression = y[, goi],
  exposure = df$Immune_Exposure,
  patch = patch_assignments
)

# Plot expression in space
ggplot(plot_df, aes(x = x, y = y, color = expression)) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  theme_void() +
  labs(title = paste("Spatial Expression of", goi),
       color = "Logcounts")
```

### Patch-level Aggregate Visualization
To better understand the spatial trends across the tissue, we can overlay the patch polygons colored by their aggregated metrics (mean expression, effect size, or t-statistic).

```{r}
#| label: patch-aggregated-plot
library(dplyr)
library(tidyr)

# Prepare aggregate data for the gene of interest (goi)
patch_metrics <- data.frame(
  patch = colnames(res$ests),
  effect_size = res$ests[goi, ],
  t_stat = res$ests[goi, ] / res$ses[goi, ],
  mean_expr = sapply(colnames(res$ests), function(p) mean(y[patch_assignments == p, goi]))
)

# Combine polygons with metrics for ggplot
# Note: 'polys' must be available (defined in 01_patches.qmd)
poly_df <- bind_rows(lapply(names(polys), function(p) {
  as.data.frame(polys[[p]]) %>%
    mutate(patch = p)
})) %>%
  left_join(patch_metrics, by = "patch")

# Plot patches colored by effect size
ggplot(poly_df, aes(x = x, y = y, group = patch, fill = effect_size)) +
  geom_polygon(color = "black", size = 0.2) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  theme_minimal() +
  labs(title = paste("Patch-level Effect Size for", goi),
       fill = "Effect Size",
       x = "x (mm)", y = "y (mm)")

# Plot patches colored by t-statistic
ggplot(poly_df, aes(x = x, y = y, group = patch, fill = t_stat)) +
  geom_polygon(color = "black", size = 0.2) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", limits = c(-5, 5), oob = scales::squish) +
  theme_minimal() +
  labs(title = paste("Patch-level T-statistic for", goi),
       fill = "T-statistic",
       x = "x (mm)", y = "y (mm)")
```
