---
title: "Spatial Stratified Differential Expression: A Case Study"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

::: callout-note
**Hackathon Proof of Concept**

This document was generated using **Gemini CLI** as part of a collaborative experimentation session during the hackathon. It represents a simple, drafty version intended as a proof of concept for the Spatial Stratified DE workflow.
:::

## Introduction

This vignette demonstrates a starting workflow for **Spatial Stratified Differential Expression (DE)**. The core idea is to define "patches" of cells that share a similar spatial context or exposure, perform DE analysis within these patches, and then aggregate the results using meta-analysis techniques.

This approach allows us to:
1.  Capture local spatial gradients that might be missed in global analyses.
2.  Account for spatial autocorrelation by modeling within contiguous units.
3.  Identify consistent biological responses across different regions of a tissue.

### Goals

-   **Step 1: Create Patches**: Build a flexible workflow to define spatial units using different inputs (e.g., cell neighborhoods, exposure hotspots).
-   **Step 2: Differential Expression**: Perform patch-wise DE while considering spatial structure.
-   **Step 3: Meta-analysis**: Summarize patch-level results to identify higher-level biological patterns and consistent gene responses.

---

## Step 1: Data Preparation and Patch Creation

We use the `MouseColonIbdCadinu2024` dataset from the `MerfishData` package, which contains MERFISH data from a mouse model of colon inflammation.

### Loading Data

```{r data-prep}
library(MerfishData)
library(SpatialExperiment)
library(InSituCor)
library(ggplot2)
library(dplyr)
library(tidyr)
library(Matrix)
library(pheatmap)
library(ggrepel)

source("patchDE.R")

# Ensure results directory exists
if (!dir.exists("../results/plots")) dir.create("../results/plots", recursive = TRUE)

# Load the dataset
spe <- MerfishData::MouseColonIbdCadinu2024()
colnames(spe) <- paste0("Cell", seq_len(ncol(spe)))

# Choose sample 5 slice 1 (3 days of DSS treatment)
spe_sub <- spe[, spe$sample_id == 5 & spe$slice_id == 1]

# Get spatial coordinates (convert to mm for patch definition)
xy <- as.matrix(spatialCoords(spe_sub))
xy_mm <- xy * 0.000109 # 1 pixel ~ 109 nm
rownames(xy_mm) <- colnames(spe_sub)
```

### Initial Tissue Exploration

Before focusing on specific cell types or neighborhoods, we visualize the entire tissue slice to understand its spatial organization. As shown in **Figure 1** and **Figure 2**, the colon displays a highly organized architecture with distinct mucosal and muscular layers. The neighborhood analysis in **Figure 2** identifies large-scale domains like `MU4`, which we will use to isolate epithelial cells in a consistent tissue context while studying local immune cell gradients.

```{r viz-initial, fig.cap="Figure 1: Tissue Slice by Cell Type (top) and Figure 2: Tissue Slice by Neighborhood (bottom)."}
# Prepare data frame for plotting
full_slice_df <- as.data.frame(xy_mm) %>%
  mutate(
    cell_type = spe_sub$tier1,
    neighborhood = spe_sub$leiden_neigh
  )

# Plot by Cell Type
p1 <- ggplot(full_slice_df, aes(x = x, y = y, color = cell_type)) +
  geom_point(size = 0.1, alpha = 0.8) +
  theme_minimal() +
  labs(title = "Tissue Slice by Cell Type",
       color = "Cell Type (Tier 1)",
       x = "x (mm)", y = "y (mm)") +
  guides(color = guide_legend(override.aes = list(size = 2)))
p1
ggsave("../results/plots/01_tissue_cell_types.png", p1, width = 8, height = 6)

# Plot by Neighborhood
p2 <- ggplot(full_slice_df, aes(x = x, y = y, color = neighborhood)) +
  geom_point(size = 0.1, alpha = 0.8) +
  theme_minimal() +
  labs(title = "Tissue Slice by Neighborhood",
       color = "Neighborhood (Leiden)",
       x = "x (mm)", y = "y (mm)") +
  guides(color = guide_legend(override.aes = list(size = 2)))
p2
ggsave("../results/plots/02_tissue_neighborhoods.png", p2, width = 8, height = 6)
```

## Defining Exposure

Our biological question is: **How do epithelial cells behave in proximity of immune cells?** 
We define the "exposure" as the number of immune cells in the local neighborhood of each epithelial cell.

```{r exposure}
# Get spatial neighbors matrix (N=50)
neighbors <- InSituCor:::nearestNeighborGraph(x = xy[,1], y = xy[,2], N = 50, subset = 1)

# Count neighboring cell types (tier1 classification)
env <- InSituCor:::neighbor_tabulate(as.vector(spe_sub$tier1), neighbors) 
rownames(env) <- rownames(xy_mm)

# Focus on Epithelial cells in a specific neighborhood (e.g., MU4)
is_epithelial <- spe_sub$tier1 == "Epithelial"
is_MU4 <- spe_sub$leiden_neigh == "MU4"
X <- env[is_epithelial & is_MU4, "Immune"]
```

### Patch Creation

We use the `getPatches` function to define spatial units optimized for studying our exposure variable.

```{r patches}
n_patches <- 100
patches <- getPatches(
    xy = xy_mm[is_epithelial & is_MU4, ],
    X = X,
    npatches = n_patches, 
    maxradius = 0.1, 
    bitesize = 0.1,
    effectivezerodist = 0.001,
    plotprogress = FALSE)

# Store patch assignments in the SpatialExperiment object
spe_sub$patch <- NA
spe_sub$patch[is_epithelial & is_MU4] <- patches
```

### Visualizing Patches

We use the `getPatches` function to define spatial units optimized for studying our exposure variable. As seen in **Figure 3**, this tiles the `MU4` neighborhood into smaller, contiguous "patches", allowing us to isolate local units that share a similar large-scale context but capture varying immune exposure.

```{r viz-patches, fig.cap="Figure 3: Epithelial Patches in MU4 neighborhood."}
# Define patch polygons for plotting
polys <- list()
use_cells_idx <- which(is_epithelial & is_MU4 & !is.na(spe_sub$patch))
for (p_id in setdiff(unique(spe_sub$patch), NA)) {
    inds <- which(spe_sub$patch[use_cells_idx] == p_id)
    if(length(inds) > 3) {
        polys[[p_id]] <- xy_mm[use_cells_idx, ][inds, ][chull(xy_mm[use_cells_idx, ][inds, ]), ]
    }
}

# Prepare df for ggplot
use_cells_df <- as.data.frame(xy_mm[use_cells_idx, ])
use_cells_df$patch <- as.factor(spe_sub$patch[use_cells_idx])

# Plot patches
p3 <- ggplot(use_cells_df, aes(x = x, y = y)) +
  geom_point(data = as.data.frame(xy_mm[is_epithelial & is_MU4, ]), aes(x = x, y = y), color = "grey80", size = 0.5) +
  geom_point(aes(color = patch), size = 0.5) +
  theme_minimal() +
  labs(title = "Epithelial Patches (MU4)", x = "x (mm)", y = "y (mm)") +
  theme(legend.position = "none") # Too many patches for a legend
p3
ggsave("../results/plots/03_epithelial_patches.png", p3, width = 8, height = 6)
```

---

## Step 2: Patch-wise Differential Expression

We now perform DE analysis within each patch using the `patchDE` function.

### Normalization

```{r normalization}
# Use logcounts from SpatialExperiment object
norm <- assay(spe_sub, "logcounts")

# Subset for epithelial cells in MU4 patches
in_patch <- !is.na(spe_sub$patch)
use <- is_epithelial & is_MU4 & in_patch

# Logcounts matrix (cells x genes) for patchDE
y <- t(as.matrix(norm[, use]))
```

### Running patchDE

```{r run-patchde}
# Define design matrix (exposure variable)
df_de <- data.frame(Immune_Exposure = env[use, "Immune"])
patch_assignments <- spe_sub$patch[use]

# Run DE across all patches
de_results <- patchDE(y = y, 
                      df = df_de, 
                      patch = patch_assignments)
```

---

## Step 3: Meta-analysis and Visualization

We combine patch-level DE estimates using inverse-variance weighting to identify genes with a consistent spatial response.

### Aggregating Results

```{r meta-agg}
# Extract results for Immune_Exposure
res <- de_results$Immune_Exposure

# Inverse-variance weighting
metawts <- pmax(res$ses, 0.1)^-1/2
metaest <- rowSums(res$ests * metawts, na.rm = TRUE) / rowSums(metawts, na.rm = TRUE)
metaSE <- sqrt(1 / rowSums(metawts))

# Calculate Z-scores and Meta p-values
metaZ <- metaest / metaSE
metaP <- 2 * pnorm(abs(metaZ), lower.tail = FALSE)
logP <- -log10(pmax(metaP, 1e-20)) # Cap for plotting

# Top consistent genes (by significance)
is_consistent <- abs(metaZ) > 3 
top_genes_names <- names(sort(abs(metaZ), decreasing = TRUE))[1:10]
```

### Global Summary (Volcano Plot)

A volcano plot identifies genes with the most consistent spatial response across all patches. As illustrated in **Figure 4**, the x-axis represents the weighted average effect size, while the y-axis shows the statistical significance (-log10 meta p-value). 

**Interpretation:** **`Egr1`** (Early Growth Response 1) emerged as a top positive hit, suggesting it may be part of an immediate-early response triggered in epithelial cells by nearby immune signals. Conversely, genes like **`Rorc`** showed negative estimates, indicating they are repressed in immune-rich areas.

```{r volcano, fig.cap="Figure 4: Meta-analysis volcano plot. Red points indicate genes with a highly consistent spatial response (|Z| > 3)."}
meta_df <- data.frame(
  gene = names(metaest),
  estimate = metaest,
  zscore = metaZ,
  logP = logP,
  label = ifelse(names(metaest) %in% top_genes_names, names(metaest), "")
)

p4 <- ggplot(meta_df, aes(x = estimate, y = logP, color = abs(zscore) > 3)) +
  geom_point(alpha = 0.4) +
  geom_text_repel(aes(label = label), size = 3, min.segment.length = 0, max.overlaps = 20) +
  theme_minimal() +
  labs(title = "Meta-analysis Volcano Plot",
       subtitle = "Top 10 genes labeled by Z-score significance",
       x = "Weighted Average Effect Size",
       y = "-log10(Meta p-value)",
       color = "|Z| > 3") +
  scale_color_manual(values = c("grey70", "firebrick")) +
  theme(legend.position = "bottom")
p4
ggsave("../results/plots/04_volcano_plot.png", p4, width = 8, height = 6)
```

### Patch-level Consistency (Heatmap)

The meta-analysis provides a global view, but we must verify that these trends hold across individual units. **Figure 5** displays the t-statistics for the top 30 genes across every individual patch.

**Interpretation:** The heatmap confirms that the response of `Egr1` and other top genes is not localized to a single "hotspot" but is fairly consistent across the majority of the sampled patches.

```{r heatmap, fig.cap="Figure 5: Heatmap of patch-level t-statistics for the top 30 consistent genes."}
# Calculate t-statistics per patch
t_stats <- res$ests / res$ses
t_stats[is.na(t_stats)] <- 0

# Select top 30 consistent genes
top_consistent_genes <- names(sort(abs(metaZ), decreasing = TRUE))[1:30]

p5 <- pheatmap(t_stats[top_consistent_genes, ], 
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-5, 5, length.out = 101),
         main = "Patch-level t-statistics (Top 30 Genes)",
         show_colnames = FALSE, silent = TRUE)
grid::grid.newpage()
grid::grid.draw(p5$gtable)
ggsave("../results/plots/05_heatmap_consistency.png", p5$gtable, width = 10, height = 8)
```

### Specific Gene Inspection: `Egr1`

The `Egr1` (Early Growth Response 1) gene is a known transcription factor involved in cellular responses to environmental stimuli.

```{r egr1-spatial, fig.cap="Figure 6: Spatial expression of Egr1 in epithelial cells (log-normalized counts)."}
goi <- "Egr1"

# Plot expression in space
plot_df <- data.frame(
  x = xy_mm[use, 1],
  y = xy_mm[use, 2],
  expression = y[, goi]
)

p6 <- ggplot(plot_df, aes(x = x, y = y, color = expression)) +
  geom_point(size = 0.5) +
  scale_color_viridis_c() +
  theme_void() +
  labs(title = paste("Spatial Expression of", goi),
       color = "Logcounts")
p6
ggsave("../results/plots/06_egr1_spatial.png", p6, width = 8, height = 6)
```

### Patch-level Aggregate Visualization

By aggregating data at the patch level in **Figure 7**, we reduce noise and can clearly visualize the "spatial effect" across the tissue architecture. 

**Interpretation:** While the global meta-analysis for `Egr1` is positive, **Figure 7** reveals local variation. Most patches are red (positive correlation), but some show a neutral or slightly blue effect, which may correspond to different crypt-villus zones or varying immune cell compositions.

```{r patch-agg-viz, fig.cap="Figure 7: Patch-level t-statistics for Egr1. Red indicates a positive correlation between Egr1 expression and immune proximity."}
# Prepare aggregate data for Egr1
patch_metrics <- data.frame(
  patch = colnames(res$ests),
  effect_size = res$ests[goi, ],
  t_stat = res$ests[goi, ] / res$ses[goi, ],
  mean_expr = sapply(colnames(res$ests), function(p) mean(y[patch_assignments == p, goi]))
)

# Combine polygons with metrics
poly_df <- bind_rows(lapply(names(polys), function(p) {
  as.data.frame(polys[[p]]) %>%
    mutate(patch = p)
})) %>%
  left_join(patch_metrics, by = "patch")

# Plot patches colored by T-statistic
p7 <- ggplot(poly_df, aes(x = x, y = y, group = patch, fill = t_stat)) +
  geom_polygon(color = "black", size = 0.2) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       limits = c(-5, 5), oob = scales::squish) +
  theme_minimal() +
  labs(title = paste("Patch-level T-statistic for", goi),
       fill = "T-statistic",
       x = "x (mm)", y = "y (mm)")
p7
ggsave("../results/plots/07_egr1_patch_tstat.png", p7, width = 8, height = 6)
```

### Conclusion
This proof-of-concept demonstrates that spatial stratification can effectively move beyond simple global correlations to identify robust, locally-validated spatial programs in complex tissues.
