---
title: "Space Mosaic"
author:
- name: Patrick Danaher
  affiliation: Bruker Spatial Biology, Seattle
  email: patrick.danaher@bruker.com
- name: Matteo Calgaro
  affiliation: Università degli Studi di Padova
  email: matteo.calgaro@unipd.it
- name: Robert Castelo
  affiliation:
    - &idupf Universitat Pompeu Fabra, Barcelona
  email: robert.castelo@upf.edu
- name: Pere Moles
  affiliation: *idupf
  email: pere.moles@upf.edu
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('SpaceMosaic')`"
vignette: >
  %\VignetteIndexEntry{Space Mosaic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeywords{GeneExpression, Spatial, Transcriptomics}
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_captions: yes
---

**License**: `r packageDescription("SpaceMosaic")[["License"]]`

```{r setup, include=FALSE}
options(width=80)
knitr::opts_chunk$set(collapse=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      comment="",
                      fig.align="center",
                      fig.wide=TRUE)
```

# Introduction

Here we illustrate how to use the SpaceMosaic package, using the Cadinu et al.
(2024) MERFIS spatial transcriptomics data available from the Bioconductor
package `r Biocpkg("MerfishData")`.

```{r load-data, message=FALSE}
library(SpatialExperiment)
library(MerfishData)

spe <- MouseColonIbdCadinu2024()
spe
```
Subset the data to sample 1 slice 1.

```{r}
spe <- spe[, spe$sample_id == "1" & spe$slice_id == "1"]
dim(spe)
```
Calculate patches that split the data into contiguous regions to maximize the
statistical power of detection.

```{r get-patches, message=FALSE}
library(SpaceMosaic)

spe <- getPatches(spe, cell_type_column="tier1",
                  response_cell_type="Epithelial",
                  explanatory_cell_type="Immune",
                  mmxpixel=0.000109,
                  npatches=40, n_iters=25)
spe
```
The patches are stored in a reduced dimension slot, named with the convention
`patches_<response_cell_type>_<explanatory_cell_type>`.

```{r}
reducedDimNames(spe)
cellpatches_epi_imm <- reducedDim(spe, "patches_Epithelial_Immune")
cellpatches_epi_imm
```
The metadata of the previous object contains patch-level information.

```{r}
metadata(cellpatches_epi_imm)
```
Visualize the patches.

```{r plot-patches, fig.width=5, fig.height=7, out.width="600px", fig.cap="Patches for Epithelial cells with respect to Immune cells."}
plotPatches(spe, "patches_Epithelial_Immune")
```
